How fluid-mechanical erosion creates anisotropic porous media

### 2022-06-27 docs/PhysicsPaper/StokesErosion.tex
NM

Like the permeability, the tortuosity provides a macroscopic perspective
of the porous medium. We define the tortuosity to be the average length
of streamlines passing through the region $[-1,1] \times [-1,1]$. In
particular, for a particular eroded geometry with velocity field $\uu$,
we calculate streamlines $\pp(t)$ that satisfy
\begin{align}
  \dot{\pp}(t) = \uu(\pp), \quad \pp(0) = (-1,y_0),
\end{align}
where $y_0 \in (-1,1)$. The streamlines are calculated with a fourth-order Runge-Kutta method. Then, the length of each streamline, which only depends on its initial $y$-coordinate, is
\begin{align}
  \lambda(y_0) = \int_{0}^{t(y_0)} \|\pp(t)\| \,dt.
\end{align}
The integral limit $t(y_0)$ is chosen so that the $x$-coordinate of
$\pp(t(y_0))$ is $+1$. Then, the longitudinal tortuosity is
\begin{align}
  T_1 = \frac{1}{2}\left(\int_{S}u(-1,y)\lambda(y)\,dy \right)
  \Bigg/
  \left(\int_{S}u(x_0,y)\,dy \right).
  \label{eqn:tortuosity_x}
\end{align} 
The factor of $1/2$ guarantees that $T_1 \geq 1$, and $T_1 = 1$ if and only if no grains are present. Identical to how we calculate the transverse permeability, we rotate each of the eroded bodies by 90 degrees and compute a transverse tortuosity $T_2$. Then, another measure of the anisotropy is
\begin{align}
  \anis = T_2/T_1.
\end{align}
As in the permeability definition of anisotropy, the random configuration of circles used to initialize the erosion simulations will have an anisotropy nearly equal to one. As this configuration erodes, the streamlines in the longitudinal direction are expected to less tortuous than those in the transverse direction, yielding $\anis > 1$.

@@

The late emergence of substantial configurational anisotropy may seem like a spurious effect that is particular to the one simulation shown in Fig.~\ref{fig3}. However, statistical analysis of many simulations will show this feature to emerge robustly in simulations with a relatively large number of bodies $M \ge 60$.

@@

>>>>

Like the permeability, the tortuosity provides a macroscopic perspective
of the porous medium. We define the tortuosity to be the average length
of streamlines passing through the region $[-1,1] \times [-1,1]$. In
particular, for a particular eroded geometry with velocity field $\uu$,
we calculate streamlines $\pp(t)$ that satisfy
\begin{align}
  \dot{\pp}(t) = \uu(\pp), \quad \pp(0) = (-1,y_0),
\end{align}
where $y_0 \in (-1,1)$. The streamlines are calculated with a fourth-order Runge-Kutta method. Then, the length of each streamline, which only depends on its initial $y$-coordinate, is
\begin{align}
  \lambda(y_0) = \int_{0}^{t(y_0)} \|\pp(t)\| \,dt.
\end{align}
The integral limit $t(y_0)$ is chosen so that the $x$-coordinate of
$\pp(t(y_0))$ is $+1$. Then, the longitudinal tortuosity is
\begin{align}
  T_1 = \frac{1}{2}\left(\int_{S}u(-1,y)\lambda(y)\,dy \right)
  \Bigg/
  \left(\int_{S}u(x_0,y)\,dy \right).
  \label{eqn:tortuosity_x}
\end{align} 
The factor of $1/2$ guarantees that $T_1 \geq 1$, and $T_1 = 1$ if and only if no grains are present. Identical to how we calculate the transverse permeability, we rotate \nick{the configuration of} eroded bodies by 90 degrees and compute a transverse tortuosity $T_2$. Then, another measure of the anisotropy is
\begin{align}
  \anis = T_2/T_1.
\end{align}
As in the permeability definition of anisotropy, the random configuration of circles used to initialize the erosion simulations will have an anisotropy nearly equal to one. As this configuration erodes, the streamlines in the longitudinal direction are expected to less tortuous than those in the transverse direction, yielding $\anis > 1$.

@@

The late emergence of substantial configurational anisotropy may seem like a spurious effect that is particular to the one simulation shown in Fig.~\ref{fig3}. However, statistical analysis of many simulations will show this feature to emerge robustly in simulations featuring a relatively large number of bodies $M \ge 60$.

@@

As a second way to quantify medium anisotropy, we measure the tortuosity, $T_1$ and $T_2$, in the longitudinal and transverse directions respectively. Figure \ref{fig4}(a), shows how both vary over the course of the simulation, with $T_2$ growing significantly larger than $T_1$, indicating that passive tracers must take more tortuous paths around the partially eroded grains when traveling transverse to the flow.
